Using Model Checking to Analyze the System Behavior of the LHC Production Grid

### 2013-02-12 ElsevierFGCS_singleColumn.tex
Introduction: added a few lines on the methodology for performing the check, and contributions. Also on the loss-of-efficiency statistics. Conclusions: added a few lines on the general applicability of the approach on other systems.

Although much effort is invested in making DIRAC reliable, entities occasionally
get into inconsistent states, leading to a loss of efficiency in both
resource usage and manpower. Debugging and fixing the root of such encountered
behaviors becomes a formidable mission due to multiple factors: the inherent
parallelism present among the system components deployed on different physical
machines, the size of the implementation ($\sim$150,000 lines of Python code),
and the distributed knowledge of different subsystems within the collaboration.

In this paper we propose the use of rigorous (formal) methods for improving
software quality. Model checking \cite{ProcessesWithData} is one such technique for analysis of an
abstract model of a system, and verification of certain system properties of
interest. Unlike conventional testing, it allows full control over the execution
of parallel processes and also supports automated exhaustive state-space
exploration. We used the mCRL2 language \cite{FormalLanguagemCRL2} and toolset \cite{mCRL2Toolset} to model the behavior of two
critical and related DIRAC components: the Workload Management (WMS) and the
Storage Management System (SMS). Based on Algebra of Communicating Processes
(ACP) \cite{process_algebra}, mCRL2 is able to deal with generic data types as well as user-defined
functions for data transformation. This makes it particularly suitable for
modeling the data manipulations made by DIRAC's agents. Visualizing the state
space and replaying scenarios with the toolkit's simulator enabled us to gain
insight into the system behavior and incrementally improve the model. 
Already using these techniques, critical race conditions and livelocks were detected and confirmed to
occur in the real system. Some of them were a result of simple coding bugs;
others unveiled more elementary design problems. We further formulated,
formalized and verified several application-specific properties.

@@

The paper is organized as follows. Section 2 introduces the architecture
of DIRAC, focusing on the two subsystems chosen as case studies. Section 3 gives a brief
overview of the mCRL2 language, and describes our approach to abstracting
and modeling the behavior of these subsystems. Section 4 presents the
analysis with the mCRL2 toolset and the issues detected. Section 5 and 6
conclude and give directions for future work, respectively.

@@

The development of DIRAC started in 2002 as a system for production of simulation
data that would serve to verify physics theory, aspects of the LHCb detector design, as
well as to optimize physics algorithms. It gradually evolved into a complete community
grid solution for data and job management, based on a general-purpose framework
that can be reused by other communities besides LHCb. Today, it covers all major
LHCb tasks starting with the raw data transfer from the experiment's detector to
the grid storage, several steps of data processing, up to the final user
analysis. 
It provides a concurrent use of over 60K CPUs and 10M file replicas distributed over
tens of grid sites, provided by the WLCG computing grid \cite{Bird:2011zz}. The community of users has grown to several hundreds, loading the 
LHC grid resources with up to 30K simultaneously running jobs (Fig.~\ref{fig:LHCb-resource-usage}, top) and over 100K jobs executed per day, during peak processing periods.
About 10PB of data is stored and visible in the LHCb file catalogs, with daily data transfers between storage units reaching 3Gbps (Fig.~\ref{fig:LHCb-resource-usage}, bottom).
\begin{figure*}[tp!]
  \centering
  {\label{fig:Job-throughput}\includegraphics[width=0.7\linewidth]{./graphics/Figure1_1.png}}    
  \vfill
  \vspace{14pt}
{\label{fig:Data-transfers}\includegraphics[width =0.7\linewidth]{./graphics/Figure1_2.png}}
  \caption{LHCb resource usage}
  \label{fig:LHCb-resource-usage}
\end{figure*}
Python was chosen as the implementation language, since it enables
rapid prototyping and development of new features. DIRAC follows the Service
Oriented Architecture (SOA) paradigm, accompanied by a network of lightweight
distributed agents which animate the system. Its main components are depicted in
Fig.~\ref{fig:DIRAC-Arch}.

@@

\section{System modeling}

\subsection{The mCRL2 language and toolset}

@@

Similar to this lightweight form of verification is static program
analysis~\cite{Schmidt98programanalysis}. Rather than code instrumentation and
execution, the objective of 
static program analysis is to determine runtime properties of systems at
compilation time.
A special analyzer targeted for the implementation language maintains
information on the control flow and function call graphs
of the software. Approximation is used by such tools in practice, in order to
improve scalability. The obvious advantage
of this approach is that it works directly on the implementation. However, most
static analysis tools aim at discovering 
more general implementation deficiencies, such as uninitialized variables, null
pointer references, or out of bound array indices. An additional drawback
is their 
practical availability for languages other than C and Java. To the best of our
knowledge, no such tool exists for Python. Both runtime verification and static analysis can be seen as additions to
traditional model checking, rather than alternatives. Some progress has been 
made in the direction of combining the
approaches~\cite{Java_PathFinder,Linux_driver,Brat01combiningstatic}. For
instance, Java PathFinder~\cite{Java_PathFinder}
relies on static analysis to reduce the state space before systematically
exploring it by executing the abstracted program.
In~\cite{Brat01combiningstatic}
static analysis and model checking also operate in a feedback loop.

@@

The investment in writing an abstract model in a formal language 
pays off proportionally to the size (number of components) of the distributed system. 
This holds especially for grid systems where a large number of concurrent
 components of the same kind deliver the overall system functionality. 
Once becoming proficient with
the formal language notation, modeling such components is
reduced to reapplying similar abstraction rules.
Practitioners
of model checking have already built sufficiently mature tools
that can be utilized (almost) as blackboxes by more traditional
software developers, hiding the mathematical theories under
the hood. Although some domain-specific knowledge is always necessary,
the descriptive guidelines we have proposed for obtaining
an abstract model can be easily 
applied to other large-scale industrial systems of a similar nature.
It goes without saying that the analyst must be careful to faithfully
describe the system behavior in the abstract model. With this critical step in mind, formal methods
are an important addition to the assembly of software analysis and verification techniques for concurrent distributed systems.
\section{Future work}
\label{sec:Section_6}
Besides discovering behavioral problems, the 
model gives opportunities to analyze the system performance in the future.
Performance analysis aims at predicting the behavior of a system with respect to dynamic properties, such as 
the expected throughput, response time, resource usage, number of requests per unit time, etc.
For this, the formal model must be enriched with timing and probability quantities attached 
to certain activities of interest.  Adding this kind of information to traditional process algebra models
means getting in the domain of stochastic process algebra. Random variables with exponential distribution are
typically associated with the designated actions, to express system transitions with duration, which could be estimates
of processing time or communication overhead. Tools based on stochastic process algebra formalisms can
then generate the underlying Markov Chain model representation and solve it to obtain steady-state probability distributions over the state space. These measures can provide a good base for assessing the system's efficiency and reliability. 
Furthermore, by varying the quantitative inputs, alternative designs can be evaluated before actual implementation
takes place. One such stochastic process algebra is PEPA (Performance Evaluation Process Algebra \cite{PEPA}), supported
by an Eclipse Plug-in for modeling and analysis of systems. 
PEPA models are annotated with information about the duration of activities and their relative probabilities of occurence.

>>>>

Although much effort is invested in making DIRAC reliable, entities occasionally
get into inconsistent states, leading to a loss of efficiency in both
resource usage and manpower. Debugging and fixing the root of such encountered
behaviors becomes a formidable mission due to multiple factors: the inherent
parallelism present among the system components deployed on different physical
machines, the size of the implementation ($\sim$150,000 lines of Python code),
and the distributed knowledge of different subsystems within the collaboration.
Depending on the frequency of certain heavier data (re)processing campaigns throughout 
the year, the occurrence of such inconsistencies can be up to several reported cases 
per month. However, since they have a blocking effect on these heavy grid activities
with a strict deadline, locating and fixing the root cause of such a problem 
typically has a lower priority within the collaboration, compared to a temporary fix, 
such as manipulating database records or restarting processes manually.

In this paper we propose the use of rigorous (formal) methods for improving
software quality. Model checking \cite{ProcessesWithData} is one such technique for analysis of an
abstract model of a system, and verification of certain system properties of
interest. Unlike conventional testing, it allows full control over the execution
of parallel processes and also supports automated exhaustive state-space
exploration. We used the mCRL2 language \cite{FormalLanguagemCRL2} and toolset \cite{mCRL2Toolset} to model the behavior of two
critical and related DIRAC components: the Workload Management (WMS) and the
Storage Management System (SMS). Based on Algebra of Communicating Processes
(ACP) \cite{process_algebra}, mCRL2 is able to deal with generic data types as well as user-defined
functions for data transformation. This makes it particularly suitable for
modeling the data manipulations made by DIRAC's agents. Visualizing the state
space and replaying scenarios with the toolkit's simulator enabled us to gain
insight into the system behavior and incrementally improve the model. 
Already using these techniques, critical race conditions and livelocks were detected and confirmed to
occur in the real system. Some of them were a result of simple coding bugs;
others unveiled more elementary design problems. We further formulated,
formalized and verified several application-specific properties. Once
the model is in place, the toolset can automatically check for general system
properties, such as deadlocks. We show how to formulate and check
application-specific
properties with the toolset. Furthermore, we outline a methodology to
effectively tackle the classical state space explosion
problem arising when the underlying model grows beyond the resources available
to perform explicit model checking. For this, we borrow 
concepts from finite state machine formalisms for description of system properties.
Extending the specification with a so-called monitoring automaton for tracking 
relevant transitions in the original model, can speed up the
analysis and discovery of traces that lead to violation of a certain property.
This idea is simple but flexible and generic enough to describe many temporal properties of concurrent programs.

@@

The paper is organized as follows. Section 2 introduces the architecture
of DIRAC, focusing on the two subsystems chosen as case studies. Section 3 gives a brief
overview of the mCRL2 language, and describes our approach to abstracting
and modeling the behavior of these subsystems. Section 4 presents the
analysis with the mCRL2 toolset and the issues detected. Work related to the 
domain of distributed systems verification is discussed in Section 5. Section 6 and 7
conclude and give directions for future work, respectively.

@@

The development of DIRAC started in 2002 as a system for production of simulation
data that would serve to verify physics theory, aspects of the LHCb detector design, as
well as to optimize physics algorithms. It gradually evolved into a complete community
grid solution for data and job management, based on a general-purpose framework
that can be reused by other communities besides LHCb. Today, it covers all major
LHCb tasks starting with the raw data transfer from the experiment's detector to
the grid storage, several steps of data processing, up to the final user
analysis. 
It provides a concurrent use of over 60K CPUs and 10M file replicas distributed
over
tens of grid sites, provided by the WLCG computing grid \cite{Bird:2011zz}. The
community of users has grown to several hundreds, loading the 
LHC grid resources with up to 30K simultaneously running jobs
(Fig.~\ref{fig:LHCb-resource-usage}, top) and over 100K jobs executed per day,
during peak processing periods.
About 10PB of data is stored and visible in the LHCb file catalogs, with daily
data transfers between storage units reaching 3Gbps
(Fig.~\ref{fig:LHCb-resource-usage}, bottom).
\begin{figure*}
  \centering
  {\label{fig:Job-throughput}\includegraphics[width=0.7\linewidth]{./graphics/Figure1_1.png}}    
  \vfill
  \vspace{14pt}
{\label{fig:Data-transfers}\includegraphics[width =0.7\linewidth]{./graphics/Figure1_2.png}}
  \caption{LHCb resource usage}
  \label{fig:LHCb-resource-usage}
\end{figure*}
Python was chosen as the implementation language, since it enables
rapid prototyping and development of new features. DIRAC follows the Service
Oriented Architecture (SOA) paradigm, accompanied by a network of lightweight
distributed agents which animate the system. Its main components are depicted in
Fig.~\ref{fig:DIRAC-Arch}.

@@

\section{System modeling}
\label{section3}
\subsection{The mCRL2 language and toolset}

@@

Similar to this lightweight form of verification is static program
analysis~\cite{Schmidt98programanalysis}. Rather than code instrumentation and
execution, the objective of 
static program analysis is to determine runtime properties of systems at
compilation time.
A special analyzer targeted for the implementation language maintains
information on the control flow and function call graphs
of the software. Approximation is used by such tools in practice, in order to
improve scalability. The obvious advantage
of this approach is that it works directly on the implementation. However, most
static analysis tools aim at discovering 
more general and low-level implementation bugs, such as uninitialized variables, null
pointer references, zero division, or out of bound array indices. 
It is almost impossible for a general static analysis tool to
figure out the exact high-level intent of the implementation code taken as input.
Because of the heavy
approximation they use,
static analysis tools are quite fast and applicable to large software projects,
but result in many false positives reported. 
An additional drawback
is their 
practical availability for languages other than C and Java. To the best of our
knowledge, no such tool exists for Python. Both runtime verification and static
analysis can be seen as additions to
traditional model checking, rather than alternatives. Some progress has been 
made in the direction of combining the
approaches~\cite{Java_PathFinder,Linux_driver,Brat01combiningstatic}. For
instance, Java PathFinder~\cite{Java_PathFinder}
relies on static analysis to reduce the state space before systematically
exploring it by executing the abstracted program.
In~\cite{Brat01combiningstatic}
static analysis and model checking also operate in a feedback loop.

@@

The investment in writing an abstract model in a formal language 
pays off proportionally to the size (number of components) of the distributed system. 
This holds especially for grid systems where a large number of concurrent
components of the same kind deliver the overall system functionality. 
It is observed that many large concurrent/distributed systems are essentially
parametrized, meaning that they contain components or process descriptions
which are behaviorally similar. Once becoming proficient with
the formal language notation, modeling such components is
reduced to reapplying similar abstraction rules. The methodology we have outlined is particularly
suitable for such systems. In addition, the guidelines proposed can be applied
when dealing with data-centered concurrent systems, where the only documentation
is the implementation itself, and most of the problems arise from sharing data among 
processes. The biggest challenge for such systems is finding a reasonable abstraction
level for the formal model, and we believe that the combination of code slicing and
abstraction rules of thumb presented in Section 3 can be helpful in this direction.

Practitioners
of model checking have already built sufficiently mature tools
that can be utilized (almost) as blackboxes by more traditional
software developers, hiding the mathematical theories under
the hood. Although some domain-specific knowledge is always necessary,
the descriptive guidelines we have proposed for obtaining
an abstract model can be easily 
applied to other large-scale industrial systems of a similar nature.
It goes without saying that the analyst must be careful to faithfully
describe the system behavior in the abstract model. With this critical step in mind, formal methods
are an important addition to the assembly of software analysis and verification techniques for concurrent distributed systems.
\section{Future work}
\label{sec:Section_6}
Besides discovering behavioral problems, the 
model gives opportunities to analyze the system performance in the future.
Performance analysis aims at predicting the behavior of a system with respect to dynamic properties, such as 
the expected throughput, response time, resource usage, number of requests per unit time, etc.
For this, the formal model must be enriched with timing and probability quantities attached 
to certain activities of interest.  Adding this kind of information to traditional process algebra models
means getting in the domain of stochastic process algebra. Random variables with exponential distribution are
typically associated with the designated actions, to express system transitions with duration, which could be estimates
of processing time or communication overhead. Tools based on stochastic process algebra formalisms can
then generate the underlying Markov Chain model representation and solve it to obtain steady-state probability distributions over the state space. These measures can provide a good base for assessing the system's efficiency and reliability. 
Furthermore, by varying the quantitative inputs, alternative designs can be evaluated before actual implementation
takes place. One such stochastic process algebra is PEPA (Performance Evaluation Process Algebra \cite{PEPA}), supported
by an Eclipse Plug-in for modeling and analysis of systems. 
PEPA models are annotated with information about the duration of activities and their relative probabilities of occurence.
